# Results
 
## Packages 
Here are the packages we use
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyverse)
library(scales)
```


## Dataset
Here are the datasets we call 
```{r}
fuel_prices_average <- read.csv("data/fuel_prices_average.csv",stringsAsFactors = FALSE)
chargStat_selected <- read.csv("data/chargStat_selected.csv", stringsAsFactors = FALSE)
wash_EV_pop_sel <- read.csv("data/wash_EV_pop_sel.csv", stringsAsFactors = FALSE)
wash_EVpop_hist_sel <- read.csv("data/wash_EVpop_hist_sel.csv", stringsAsFactors = FALSE)
```


## Result Guideline:
This analysis explores the electric vehicle (EV) market and its adoption trend, including fuel prices, charger distribution, and state-level EV sales. By examining trends, features, and key factors impacting EV adoption, we aim to understand its ecosystem. A focused analysis on Washington State will highlight specific EV sales types, charger availability, and brand performance.



## Fuel Prices Time Series Trend
This is the code to preprocess for the coming graph
```{r, warning=FALSE}
fuel_prices_average <- fuel_prices_average |>
  mutate(across(
    c(Gasoline, E85, CNG, LNG, Propane., Diesel, B20, B99.B100),
    ~ as.numeric(gsub("[^0-9.]", "", .))  # Remove non-numeric characters like '$'
  ))


na_summary <- sapply(fuel_prices_average[c("Gasoline", "E85", "CNG", "LNG", "Propane.", "Diesel", "B20", "B99.B100")], function(x) sum(is.na(x)))

# Convert Report.Date to Date format
fuel_prices_average <- fuel_prices_average %>%
  mutate(Report.Date = as.Date(Report.Date, format = "%m/%d/%y"))

```

In the interest of understanding EV adoption, analyzing fuel prices is crucial as they serve as a significant substitute factor for EV usage. This graph examines the trends in Gasoline, E85, CNG, Propane, and Diesel prices from 2001 to 2024, collected at intervals of 4-6 months. These fuel types were chosen due to their relevance in the automotive industry: Gasoline and Diesel are dominant in traditional vehicles, E85 represents a renewable alternative for flex-fuel cars, and CNG and Propane cater to niche or commercial uses. The graph reveals significant insights. First, Gasoline and Diesel prices show correlated trends, with spikes during periods of economic disruption. Second, E85 and CNG display relative price stability, likely reflecting consistent demand and supply patterns. Lastly, Propane prices exhibit sharp seasonal volatility, highlighting the influence of weather and supply chain factors. Overall, these trends underscore how fuel price dynamics can impact consumer decisions regarding EV adoption, presenting EVs as a cost-stable alternative amidst fuel price volatility.

```{r,warning=FALSE}


ggplot(fuel_prices_average, aes(x = Report.Date)) +
  geom_line(aes(y = Gasoline, color = "Gasoline"), linewidth = 0.5) +
  geom_line(aes(y = E85, color = "E85"), linewidth = 0.5) +
  geom_line(aes(y = CNG, color = "CNG"), linewidth = 0.5) +
  geom_line(aes(y = Propane., color = "Propane"), linewidth = 0.5) +
  geom_line(aes(y = Diesel, color = "Diesel"), linewidth = 0.5) +
  labs(
    title = "Time Series of Gasolino, E85, CNG, Propane, and Diesel",
    x = "Date",
    y = "Price (USD)",
    color = "Fuel Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
## 2-year Moving Average Fuel Price Trend
The 2-year moving average smooths fuel price trends (2001–2024), revealing long-term stability for Propane and Gasoline, with Diesel and E85 showing higher volatility. These trends highlight fuel cost dynamics influencing EV adoption.



```{r}


fuel_prices_MA <- fuel_prices_average |>
  mutate(
    Gasoline_MA = zoo::rollmean(Gasoline, k = 4, fill = NA, align = "right"),
    E85_MA = zoo::rollmean(E85, k = 4, fill = NA, align = "right"),
    CNG_MA = zoo::rollmean(CNG, k = 4, fill = NA, align = "right"),
    Propane_MA = zoo::rollmean(Propane., k = 4, fill = NA, align = "right"),
    Diesel_MA = zoo::rollmean(Diesel, k = 4, fill = NA, align = "right")
  )


fuel_prices_MA_filtered <- fuel_prices_MA |>
  filter(!is.na(Gasoline_MA) | !is.na(E85_MA) | !is.na(CNG_MA) | 
         !is.na(Propane_MA) | !is.na(Diesel_MA))

ggplot(fuel_prices_MA_filtered, aes(x = Report.Date)) +
  geom_line(aes(y = Gasoline_MA, color = "Gasoline"), linewidth = 1,linetype = "dashed") +
  geom_line(aes(y = E85_MA, color = "E85"), linewidth = 1,linetype = "dashed") +
  geom_line(aes(y = CNG_MA, color = "CNG"), linewidth = 1,linetype = "dashed") +
  geom_line(aes(y = Propane_MA, color = "Propane"), linewidth = 1,linetype = "dashed") +
  geom_line(aes(y = Diesel_MA, color = "Diesel"), linewidth = 1,linetype = "dashed") +
  labs(
    title = "2 years Moving Average of Fuel Prices (Filtered)",
    x = "Date",
    y = "Price (USD)",
    color = "Fuel Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

## Electric Vehicle Chargers Station (In Fuel Station)
After observing the rising and fluctuating trends in fuel prices, we now turn our attention to the distribution of electric vehicle (EV) charging stations, which serve as the “fueling” infrastructure for EVs. This analysis highlights that California (CA) leads with the highest number of EV chargers, significantly outpacing other states. The next top 10 states with notable EV charger distributions are New York (NY), Texas (TX), Connecticut (CT), Michigan (MI), Florida (FL), Ohio (OH), Pennsylvania (PA), Massachusetts (MA), Illinois (IL), and Washington (WA). This distribution underscores the regional dynamics of EV infrastructure development.


```{r}
state_distribution <- chargStat_selected |>
  group_by(State) |>
  summarise(Count = n()) |>
  mutate(Percentage = Count / sum(Count) * 100) |>
  arrange(desc(Percentage))


ggplot(state_distribution, aes(x = Percentage, y = reorder(State, Percentage))) +
  geom_point(aes(color = Percentage), size = 1) +
  labs(
    title = "Percentage of EV Chargers in Fuel Stations by State",
    x = "Percentage (%)",
    y = "State"
  ) +
  scale_color_gradient(low = "lightblue", high = "darkblue", name = "Ranking") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    legend.position = "right"
  ) 
```

## EV Charge Station By Count 
EV Level 1 chargers are slow and mainly used in residential settings for overnight charging, hence there is not too much device in public area. EV Level 2 chargers provide faster charging, suitable for public and commercial use. DC Fast Chargers offer rapid charging for long-distance travel. The graphs display the distribution of these charger types across states. Level 1 chargers are sparse, reflecting their limited usage. Level 2 chargers dominate the landscape, being widely used in public and private spaces. DC Fast Chargers, crucial for quick charging, are concentrated in states like California and Texas. These charts highlight California’s leadership in EV infrastructure, with other states trailing significantly.
```{r}
charger_type_distribution <- chargStat_selected |>
  group_by(State) |>
  summarise(
    Level1_Chargers = sum(EV.Level1.EVSE.Num, na.rm = TRUE),
    Level2_Chargers = sum(EV.Level2.EVSE.Num, na.rm = TRUE),
    DC_Fast_Chargers = sum(EV.DC.Fast.Count, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = c(Level1_Chargers, Level2_Chargers, DC_Fast_Chargers),
    names_to = "Charger.Type",
    values_to = "Total_Count"
  ) |>
  arrange(desc(Total_Count))


charger_level1 <- charger_type_distribution |>
  filter(Charger.Type == "Level1_Chargers") |>
  arrange(desc(Total_Count))

charger_level2 <- charger_type_distribution |>
  filter(Charger.Type == "Level2_Chargers") |>
  arrange(desc(Total_Count))

charger_dc_fast <- charger_type_distribution |>
  filter(Charger.Type == "DC_Fast_Chargers") |>
  arrange(desc(Total_Count))


ggplot(charger_level1, aes(x = reorder(State, -Total_Count), y = Total_Count, fill = Total_Count)) +
  geom_bar(stat = "identity", color = "white") +
  labs(
    title = "EV Level1 EVSE Num by State",
    x = "State",
    y = "Total Count"
  ) +
  scale_fill_viridis_c(option = "D", direction = -1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


ggplot(charger_level2, aes(x = reorder(State, -Total_Count), y = Total_Count, fill = Total_Count)) +
  geom_bar(stat = "identity", color = "white") +
  labs(
    title = "EV Level2 EVSE Num by State",
    x = "State",
    y = "Total Count"
  ) +
  scale_fill_viridis_c(option = "D", direction = -1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


ggplot(charger_dc_fast, aes(x = reorder(State, -Total_Count), y = Total_Count, fill = Total_Count)) +
  geom_bar(stat = "identity", color = "white") +
  labs(
    title = "EV DC Fast Chargers by State",
    x = "State",
    y = "Total Count"
  ) +
  scale_fill_viridis_c(option = "D", direction = -1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


Our focus is on public chargers, and the following graph provides a detailed comparison between Level 2 chargers and DC Fast Chargers across states. Level 2 chargers, typically used for residential and commercial purposes, offer a moderate charging speed, making them suitable for longer parking durations. On the other hand, DC Fast Chargers are designed for rapid charging, catering to drivers on long-distance trips or in need of quick top-ups. From the graph, we observe that states with a higher total number of EV chargers, such as California, Texas, and New York, also show a significant presence of both charger types, reflecting a balanced and comprehensive infrastructure.

```{r}
level2_and_fast_distribution <- chargStat_selected |>
  select(State, EV.Level2.EVSE.Num, EV.DC.Fast.Count) |>
  pivot_longer(
    cols = c(EV.Level2.EVSE.Num, EV.DC.Fast.Count),
    names_to = "Charger.Type",
    values_to = "Count"
  ) |>
  group_by(State, Charger.Type) |>
  summarise(Total_Count = sum(Count), .groups = "drop") |>
  arrange(Charger.Type, desc(Total_Count))

# Create faceted bar plot
ggplot(level2_and_fast_distribution, aes(x = reorder(State, -Total_Count), y = Total_Count, fill = Charger.Type)) +
  geom_bar(stat = "identity", color = "white", position = "stack") +
  labs(
    title = "Level 2 and DC Fast Charger Distribution by State",
    x = "State",
    y = "Total Count",
    fill = "Charger Type"
  ) +
  scale_fill_viridis_d(option = "D", direction = -1) + # Adjust for distinct color for Charger Type
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
This distribution highlights a clear correlation between the overall availability of chargers and the diversity of charging options provided. States with larger EV adoption rates prioritize offering a mix of Level 2 and DC Fast Chargers to cater to various consumer needs. Interestingly, states with fewer chargers, such as Wyoming and Alaska, exhibit limited diversity, focusing primarily on either Level 2 or DC Fast Chargers. This trend underscores how states with robust EV ecosystems invest in accessibility and convenience, promoting EV adoption by addressing both daily and long-distance charging requirements. 

## Charge Station Access ability 

This pie chart reveals that a significant majority, approximately 66.9%, of fuel stations equipped with EV chargers operate 24 hours a day, ensuring round-the-clock accessibility for EV owners. This highlights the growing emphasis on convenience and accessibility in the EV charging infrastructure, catering to diverse consumer needs such as late-night charging or emergency top-ups. Conversely, around 33.1% of these stations do not operate 24 hours, which may reflect limitations in certain locations, such as rural areas, or operational constraints like staffing or energy costs. This disparity underscores the need for further infrastructure development, particularly in states where 24-hour access is limited, to support the expanding EV market and provide equitable access to charging facilities.
```{r}
access_time_grouped <- chargStat_selected |>
  mutate(
    Access_Group = ifelse(grepl("^24 hours daily", Access.Days.Time), "24 Hours", "non 24 hours")
  )


overall_distribution <- access_time_grouped |>
  group_by(Access_Group) |>
  summarise(Count = n(), .groups = "drop") |>
  mutate(Percentage = (Count / sum(Count)) * 100)


ggplot(overall_distribution, aes(x = "", y = Percentage, fill = Access_Group)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 5, color = "white") +
  labs(
    title = "Distribution of EV Chargers Access Time (U.S)",
    fill = "Access Group"
  ) +
  theme_void() +
  scale_fill_manual(values = c("24 Hours" = "dodgerblue", "non 24 hours" = "tomato"))

```


In vary by state, we notice that the majority of EV charger stations operate 24 hours, as seen in states like California and New York, where the percentage is consistently high. However, states with lower overall counts, such as New Mexico and Alaska, show a relatively higher proportion of non-24-hour stations. This observation builds on our earlier exploration of state-level charger distribution, emphasizing that states with more extensive EV infrastructure also tend to prioritize 24-hour accessibility. The graph highlights regional differences, reflecting how infrastructure and accessibility priorities align with state-level EV adoption and public charging needs.

```{r}


state_access_distribution <- access_time_grouped |> 
  group_by(State, Access_Group) |> 
  summarise(Count = n(), .groups = "drop") |> 
  group_by(State) |> 
  mutate(Percentage = (Count / sum(Count)) * 100)


state_order <- state_access_distribution |>
  filter(Access_Group == "24 Hours") |>
  arrange(desc(Percentage)) |>
  pull(State)

ggplot(state_access_distribution, aes(x = factor(State, levels = state_order), y = Percentage, fill = Access_Group)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  labs(
    title = "Distribution of Access Times in EV Charger Stations (24 Hours Ranked)",
    x = "State",
    y = "Percentage (%)",
    fill = "Access Group"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  scale_fill_manual(values = c("24 Hours" = "green", "non 24 hours" = "purple"))
```
## Proportion of EVs in Cities in WA
After examining the overall state of EV chargers across the United States, we narrowed our focus to Washington State (WA) to analyze how EVs and charging stations are distributed across its counties. To start, we created a bar graph showing the proportion of EVs in each county within WA.

```{r}
wash_city_EVs <- wash_EV_pop_sel |>
  group_by(City) |>
  summarise(CarCount = n()) |>
  arrange(desc(CarCount)) 
```

```{r}
#| fig-width: 25
#| fig-height: 25
wash_city_EVs2 <- wash_city_EVs |>
  mutate(City = ifelse(CarCount < 500, "Other", as.character(City))) |>
  group_by(City) |>
  summarise(CarCount = sum(CarCount)) |>
  mutate(Proportion = CarCount / sum(CarCount)) |>
  arrange(desc(Proportion))

ggplot(wash_city_EVs2, aes(y = reorder(City, -Proportion), x = Proportion, fill = Proportion)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Proportion of Electric Vehicles in Each City",
    x = "Proportion",
    y = "City"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15),
    axis.text.y = element_text(size = 15),
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18)
  ) +
  scale_x_continuous(labels = percent) +  # Show x-axis as percentages
  scale_fill_gradient(low = "lightblue", high = "darkblue")  # Gradient color scheme
```
In this graph, cities with fewer than 500 EVs were grouped into a single category labeled as “Other.” 

The city of Seattle stands out with the highest proportion of EVs, followed by Bellevue and Vancouver. Conversely, many smaller cities, such as Spanaway, Stanwood, and Wenatchee, show significantly lower proportions of EVs. This demonstrates a skewed distribution, where a few cities dominate EV ownership while the majority contribute minimally.

This result prompts further exploration of the factors that may correlate with EV proportions or counts, such as: availability of charging infrastructure, population density, Income levels, environmental policies or incentives.


## Number of EVs and Charging Stations in Cities in WA

Using our dataset, we aim to investigate whether the number of charging stations influences EV ownership levels.


```{r}
wash_chargStat <- chargStat_selected |>
  filter(State == "WA", Restricted.Access == "false")

wash_charg_counts <- wash_chargStat |>
  group_by(City) |>
  summarise(StationCount = n()) |>
  arrange(desc(StationCount)) 

merged_data <- merge(wash_city_EVs, wash_charg_counts, by = "City", suffixes = c("car_", "station_"))

ev_threshold <- 10000
station_threshold <- 10

ggplot(merged_data, aes(y = CarCount, x = StationCount)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  geom_text(
    aes(label = ifelse(CarCount > ev_threshold | StationCount > station_threshold, City, "")), 
    hjust = 0.5, vjust = -0.5, size = 3, color = "darkred"
  ) +
  theme_minimal() +
  labs(
    title = "Scatterplot of Station Counts and EV Counts in Cities in WA",
    y = "Counts of EVs in Each City",
    x = "Counts of Stations in Each City"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```
This scatterplot illustrates the relationship between the number of EV charging stations (x-axis) and the number of EVs (y-axis) in various cities in Washington State (WA). The cities with high number of EVs(more than 10,000) and relatively high numbers of charging stations(more than 10) are labeled in the graph. 

Seattle stands out with both the highest number of EVs (over 30,000) and a relatively high number of charging stations. Bellevue is another prominent city, with over 10,000 EVs but fewer charging stations compared to the other cities. Tacoma and Spokane are notable for having much lower EV counts but relatively more charging stations. 

The graph shows that charging stations alone do not fully explain EV adoption. Other factors, such as population size, income levels, environmental incentives, and commuting patterns, likely contribute more. 

In addition, we can hypothesize that there might be a threshold for the number of charging stations to significantly influence EV adoption. Specifically, if the number of charging stations exceeds 25 (as seen in Seattle), it could lead to a substantial increase in the number of EVs. This can be tested in the later studies with more data of other States. 


## Distributions of EV Types in WA

We also want to understand why the number of EVs in some cities does not strongly correlate with the number of charging stations. This is particularly evident in cities like Bellevue, where there are relatively few charging stations but a high number of EVs. One possible explanation is that residents may be purchasing more Plug-in Hybrid Electric Vehicles (PHEVs) rather than Battery Electric Vehicles (BEVs). To explore this, we began by examining the overall distribution of these two types of EVs.

```{r}
ev_type_counts <- wash_EV_pop_sel |>
  count(Electric.Vehicle.Type) |>
  mutate(percentage = n / sum(n) * 100)

ggplot(ev_type_counts, aes(x = "", y = percentage, fill = Electric.Vehicle.Type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(
    title = "Distribution of EV Types",
    fill = "EV Type"
  ) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5))
```
This pie chart illustrates the distribution of Electric Vehicle (EV) types in WA, comparing Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs). 

BEVs make up 78.8% of the total EV population, showing a strong preference for fully electric vehicles over hybrids. PHEVs account for 21.2% of the total EV population. This suggests that the majority of EV users are choosing vehicles that rely entirely on electricity rather than hybrids that use a combination of fuel and electricity. 

This general distribution pie chart alone is insufficient to explain why some cities with few charging stations have a high number of EVs, while others with relatively more charging stations have fewer EVs. To better understand this, we need to analyze the relationship between the proportion of BEVs and the number of charging stations in each city to identify potential patterns or trends.

## Number of Charging Stations and Distributions of EV Types in Cities in WA

```{r}
bev_proportions <- wash_EV_pop_sel |> 
  group_by(City) |>
  summarise(
    Total_EV_Count = n(),
    BEV_Count = sum(Electric.Vehicle.Type == "Battery Electric Vehicle (BEV)"),
    BEV_Proportion = BEV_Count / Total_EV_Count,
    .groups = "drop"
  )

scatter_data <- merged_data |>
  select(City, StationCount) |>
  inner_join(bev_proportions, by = "City")

ggplot(scatter_data, aes(x = StationCount, y = BEV_Proportion)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Scatterplot of BEV Proportion vs Stations in Cities",
    x = "Number of Stations in Each City",
    y = "Proportion of BEVs in Each City"
  ) +
  scale_y_continuous(labels = scales::percent) +  
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```
This scatterplot illustrates the relationship between the proportion of Battery Electric Vehicles (BEVs) (y-axis) and the number of charging stations in each city (x-axis). 

Many cities have a BEV proportion above 75%, regardless of the number of charging stations, indicating a strong preference for BEVs across various regions. There are also a few cities with very low BEV proportions (below 25%), but these are less frequent. The majority of cities have fewer than 5 charging stations, yet their BEV proportions remain relatively high. This suggests that BEV adoption is not entirely dependent on public charging infrastructure. This conclusion is supported by the next graph, which focuses on the distribution of EV types in cities with a high number of EVs.  

```{r}
ev_threshold <- 5000

high_cities <- merged_data |>
  filter(CarCount > ev_threshold) |> 
  select(City, StationCount)

high_cities_ev <- wash_EV_pop_sel |>
  filter(City %in% high_cities$City)

wash_city_ev_prop <- high_cities_ev |>
  group_by(City, Electric.Vehicle.Type)  |>
  summarise(Count = n(), .groups = "drop")  |>
  group_by(City)  |>
  mutate(Proportion = Count / sum(Count)) |>
  left_join(high_cities, by = "City") |>
  arrange(StationCount) |>
  mutate(City = factor(City, levels = unique(City)),
    Electric.Vehicle.Type = factor(Electric.Vehicle.Type, levels = c(
    "Plug-in Hybrid Electric Vehicle (PHEV)",
    "Battery Electric Vehicle (BEV)"
  )))

ggplot(wash_city_ev_prop, aes(x = reorder(City, StationCount), y = Proportion, fill = Electric.Vehicle.Type)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(
    title = "Proportion of Electric Vehicle Types in Cities with High Counts",
    x = "City",
    y = "Proportion",
    fill = "EV Type"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )
```
This stacked bar chart depicts the proportions of Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) in cities with high EV counts(more than 5000), and the the cities on x-axis are ordered by the increasing number of charging stations in that city.

Across all cities shown (Bothell, Bellevue, Kirkland, Sammamish, Redmond, Olympia, Renton, Vancouver, and Seattle), BEVs consistently represent the majority of EVs. While BEVs dominate, the proportion of PHEVs varies slightly among cities. Cities like Olympia and Seattle show slightly higher proportions of PHEVs compared to others. Kirkland, Sammamish, and Redmond have lower PHEV proportions, with BEVs accounting for almost the entire EV population. However, there is not patter of the number of BEVs from left bar to the right bar, meaning from a relatively low number of charging stations to a high number charging stations. 

This emphasizes the previous conclusion that the reason why the number of EVs is not correlated with the number of charging stations is not related to the type of EVs.  


## Distribution of EV Makes in WA

Using the data, we also want to know how the make of EVs are disrtibuted. Are they dominated by one make?

```{r}
make_counts <- wash_EV_pop_sel|>
  count(Make) |>
  arrange(desc(n)) |>
  mutate(Make = factor(Make, levels = Make)) 

threshold <- 500
make_counts <- make_counts |>
  mutate(Make = ifelse(n < threshold, "OTHER", as.character(Make))) |>
  group_by(Make) |>
  summarise(n = sum(n)) |>
  arrange(desc(n)) |>
  mutate(Make = factor(Make, levels = Make))

ggplot(make_counts, aes(x = Make, y = n, fill = n)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
  theme_minimal() +
  labs(
    title = "Distribution of Vehicle Makes",
    x = "Vehicle Make",
    y = "Count",
    fill = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```
This graph showcases the distribution of total EV sales by brand in Washington State. Tesla emerges as the clear leader, significantly outpacing other manufacturers, reflecting its dominance in the EV market. Chevrolet and Nissan hold the second and third positions, respectively, but their sales are dwarfed by Tesla’s overwhelming presence. Other brands, such as Ford, Kia, and BMW, show modest contributions to the market, while smaller brands trail even further behind. This distribution underscores Tesla’s pivotal role in driving EV adoption in the state, with other brands collectively contributing to a diverse yet secondary share of the EV market.
